---
title: "Displaying data using graphs and tables"
author:

<!--  output: html_document -->
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

<!--
knitr:: seems to be the best way to control figure dimensions with ggplots. 
It may be necessary to force a different width and height for individual plots 
in base R, e.g., see mosaicplot() and plot() used for scatter plot of guppy data
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5.5, fig.height=5)
```

You can draw plots using commands in base R or use ```ggplot()``` from the tidyverse. Each option has passionate adherents and it is pointless to argue. Try both and see which you find most appealing and easiest to use. If you need help, check the "Graphs & Tables" tab at the [R tips](https://www.zoology.ubc.ca/~schluter/R/) page.

***

## Data set 1: Mammal body mass
These data were published as a data paper in Ecology and deposited in the <i>Ecological Archives</i> (Smith et al. 2003. Body mass of late Quaternary mammals. <i>Ecology</i> 84: 3403.) See the [metadata](http://www.esapubs.org/archive/ecol/E084/094/) for a description.

Most of the variables are categorical, with multiple named categories. ```continent``` includes mammals on islands ("Insular" category) whereas "Oceanic" refers to marine mammals. Body mass (in grams) is the sole numeric variable. The ```status``` variable indicates whether species is currently present in the wild (extant), extinct as of late Pleistocene (extinct), extinct within the last 300 years (historical), or an introduced species (introduction).

<br>

### Read and examine the data

The original data were saved in mammals.csv file on our server [here](https://www.zoology.ubc.ca/~bio501/R/data/mammals.csv). Download the file to your computer and open in a spreadsheet program (e.g., Excel, Calc) to have a look at it.

Start R and read the contents of the file to a data frame. 

Use the <code>head()</code> function to view the first few lines of the data frame on the screen. You'll see that every row represents the data for a different mammal species.

<br>

### Frequency tables

1. Which continent has the greatest number of mammal species? Which has the least? Make a table of the frequency of cases on each continent (remember that the category "NA" in continent stands for North America, not missing data).

2. You'll notice that one of the continents is missing - North America. Yet if you open the spreadsheet you'll see that it is in the data set. Can you figure out why it is missing? Answer is at the bottom of this page*, but see if you can figure it out yourself first. (Don't waste too much time on it.)

3. You'll also notice in the frequency table for the variable ```continent``` that there's a typo in the data. One case is shown as having the continent "Af" rather than "AF for Africa". Fix this using the command line in R and recalculate the frequency table.

4. How many extinct mammals are recorded in the data file? Use a frequency table to find out.

5. Create a two-way frequency table (contingency table) showing the status of mammal species on each continent. 

6. It is easier to appreciate the relative number of extinctions on the continents if you also include the marginal sums in the table (to give the total number of species). Without doing the calculations, you can probably see which continent has the greatest number of extinctions relative to the number of extant species. Which continent?

[//]: <!--
### Answers
All lines below beginning with double hashes are R output

```{r eval = TRUE, message = FALSE}
# Load the packages you might need
library(tidyverse)

# Read and inspect the data
mammals <- read.csv(url("https://www.zoology.ubc.ca/~bio501/R/data/mammals.csv"))
head(mammals)

```

1. Number of mammal species on each continent.

```{r eval = TRUE, message = FALSE}
table(mammals$continent)
```

2. Fix the problem of missing North America (unfortunately coded as NA in the continent column). 

```{r eval = FALSE, message = FALSE}
mammals <- read.csv(url("https://www.zoology.ubc.ca/~bio501/R/data/mammals.csv"), 
                na.strings = "")

# tidyr version
mammals <- read_csv(url("https://www.zoology.ubc.ca/~bio501/R/data/mammals.csv"), 
                na = c(""))
```

3. Fix "Af" typo.

```{r eval = TRUE, message = FALSE}
which(mammals$continent=="Af")
mammals$continent[322]<-"AF"
table(mammals$continent)
```

4. How many extinct mammals? The table shows that 242 species of mammal are listed as extinct. 

```{r eval = TRUE, message = FALSE}
z <- table(mammals$status)
z
```

5. Extinction status by continent (contingency table). Australia (AUS) seems to have the greatest number of extinct species relative to the total number. This is easier to appreciate if the row sums are included in the table. 

```{r eval = TRUE, message = FALSE}
# Contingency table
table(mammals$continent, mammals$status)

# Include row sums using base R
mytab <- table(mammals$continent, mammals$status)
addmargins(mytab, margin = c(1,2), FUN = sum, quiet = TRUE)

# Include row sums using tidyverse. The last mutate command is optional and converts NA to zero.
mytab <- summarize(group_by(mammals, continent, status), n = n())
mytab <- spread(mytab, status, n)
mytab
mytab <- mutate(mytab, across(everything(), ~replace_na(., 0)))
as.data.frame(mytab)

```

[//]: -->

<br>

### Graphing frequency distributions

1. Plot the number of mammal species on each continent using a simple bar graph. Include a label for the y axis.

2. The plot categories are listed in alphabetical order by default, which is arbitrary and makes the visual display less efficient than other possibilities. What is the most sensible order for the categories? Redo the bar graph using this order.

3. Generate a histogram of the body masses of mammal species. How informative is that?!

4. Create a new variable in the mammal data frame: the log (base 10) of body mass. (See "Transform" on the R tips "Data" page if you need help with this.)

5. Generate a histogram of log body mass. Is this more informative? Morphological data commonly require a log-transformation to analyze.

6. Redo the previous histogram but use a bin width of 2 units. How much detail is lost? Redo the histogram but try a bin width of of 1; then try 0.5; and then 0.1. Which bin width is superior?

7. Redo the histogram, but display probability density instead of frequency.

8. How does the frequency distribution of log body mass depart from a normal distribution? Answer by visual examination of the histogram you just created. Now answer by examining a normal quantile plot instead. Which display is more informative? Do the data conform to a normal distribution?

9. Optional: redraw the histogram of log body mass and superimpose a normal density curve to help visualize deviations from normality. In what ways do the data depart from normality?

[//]: <!--

### Answers

```{r eval = TRUE, message = FALSE}
# 1. Bar plot of mammal species by continent
barplot(table(mammals$continent), col="firebrick", cex.names=0.8, 
        ylim=c(0,1600), las = 1)

# 2. It is best to order the categories in order of decreasing frequency.
barplot(sort(table(mammals$continent), decreasing=TRUE), col="firebrick", 	
        cex.names=0.8, las = 1, ylim=c(0,1600), ylab="Frequency")
```

Alternatively, use ```ggplot()``` instead of base R.

```{r eval = TRUE, fig.width = 5, fig.height=4}
library(ggplot2)
ggplot(mammals, aes(x = continent)) + 
	geom_bar(stat = "count", fill = "firebrick") +
	labs(x = "Continent", y = "Frequency") +
  theme_classic()

# To order by category in ggplot, first make a new factor variable
mammals$continent_ordered <- factor(mammals$continent, 
				levels = names(sort(table(mammals$continent), decreasing = TRUE)) )

ggplot(mammals, aes(x = continent_ordered)) + 
	geom_bar(stat = "count", fill = "firebrick") +
	labs(x = "Continent", y = "Frequency") +
  theme_classic()
```

Histogram of body masses. Results with different bin widths not shown.

```{r eval = TRUE}
# 3. Histogram
hist(mammals$mass.grams, col="firebrick", right = FALSE, las = 1, 
     xlab = "Body mass (g)", main = "")
```
```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(mammals, aes(x = mass.grams)) + 
	geom_histogram(fill = "firebrick", col = "black", boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()
```

```{r eval = TRUE}
# 4. Add a new variable, log10 of body mass
mammals$logmass <- log10(mammals$mass.grams)
```

```{r eval = TRUE}
# 5. Histogram
hist(mammals$logmass, col="firebrick", right = FALSE, las = 1, 
     xlab = "Log10 body mass", main = "")
```

or

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", boundary = 0, binwidth = 0.5) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()
```

```{r eval = TRUE, fig.width = 5, fig.height=4}
# 6. Vary bin width
ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 2, boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()

ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 1, boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()

ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 0.5, boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()

ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 0.1, boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()
```


```{r eval = TRUE}
# 6. Change bin width to 2
hist(mammals$logmass, col="firebrick", right = FALSE, las = 1, 
     xlab = "Log10 body mass", main = "", breaks = seq(0, 10, by = 2))
```

or

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 2, boundary = 0) + 
	labs(x = "log10 body mass", y = "Frequency") + 
	theme_classic()
```

7. Plot density instead

```{r eval = TRUE}
hist(mammals$logmass, col = "firebrick", right = FALSE, las = 1, prob = TRUE,
     xlab = "Log10 body mass", main = "", breaks = seq(0, 8.5, by = 0.5))
```

or

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "firebrick", col = "black", binwidth = 0.5, 
	             boundary = 0, aes(y = after_stat(density))) + 
	labs(x = "log10 body mass", y = "Density") + 
  theme_classic()

```

8. Normal quantile plot.

```{r eval = TRUE}
qqnorm(mammals$logmass)
qqline(mammals$logmass) # adds the straight line for comparison through 1st and 3rd quartiles
```

9. Histogram with best-fit normal curve superimposed.

```{r eval = TRUE}
# The curve function is fussy about the name of the variable: must be "x"
x <- mammals$logmass
hist(x, col="firebrick", right = FALSE, las = 1, prob = TRUE,
     xlab = "Log10 body mass", main = "", breaks = seq(0, 8.5, by = 0.5))
m <- mean(x, na.rm = TRUE)
s <- sd(x, na.rm = TRUE)
curve(dnorm(x, mean = m, sd = s), col="red", lwd = 2, add = TRUE)

```

[//]: -->

<br>

### Comparing frequency distributions

1. Use a box plot to compare the distribution of body sizes (log scale most revealing) of mammals having different extinction status. Are extinct mammals similar to, larger than, or smaller than, extant mammals?

2. Examine the previous box plot. How do the shapes of the body size distributions compare between extinct and extant mammals?

3. Redo the previous box plot but make box width proportional to the square root of sample size. Add a title to the plot.

4. Optional: Draw a violin plot to compare the frequency distribution of log body sizes of mammals having different extinction status. Which do you find is more revealing about the shapes of the body size distributions: box plot or violin plot?

5. Use multiple histograms to compare the frequency distribution of log body sizes of mammals having different extinction status. Stack the panels one above the other. In this plot, how easy is it to visualize differences among treatments in the distributions compared to your previous plots?

6. Make a table of the median log body mass of each extinction-status group of mammals. Are the values consistent with the plotted distributions?

[//]: <!--
### Answers

The graphs show that extinct mammals tend to have large mass compared to extant mammals. The frequency distributions for these two groups also have opposite skew, with extinct mammals having left skew. 

```{r eval = TRUE}
# 1. Box plot to compare the distribution of body sizes
boxplot(logmass ~ status, data = mammals, ylab = "log10 body mass", 
        col = "goldenrod1", las = 1)
```

or 

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(mammals, aes(x = status, y = logmass)) + 
	geom_boxplot(fill = "goldenrod1", notch = FALSE) + 
	labs(x = "Status", y = "Log10 body mass") + 
	theme_classic()
```

```{r eval = TRUE}
# 3. Box width proportional to the square root of sample size
boxplot(logmass ~ status, data = mammals, ylab = "log10 body mass", 
        col = "goldenrod1", las = 1, varwidth = TRUE, 
        main = "Body sizes of mammals by status")
```

```{r eval = TRUE}
# 4. Violin plot
ggplot(mammals, aes(x = status, y = logmass)) + 
	geom_violin(fill = "goldenrod1") + 
	labs(x = "Status", y = "Log10 body mass") + 
	stat_summary(fun = mean,  geom = "point", color = "black") +
  theme_classic()
```

```{r eval = TRUE, fig.width = 5, fig.height=6}
# 5. Multiple histograms
ggplot(mammals, aes(x = logmass)) + 
	geom_histogram(fill = "goldenrod1", col = "black", 
	         binwidth = 0.2, boundary = 0) +
	labs(x = "log10 body mass", y = "Frequency") + 
	facet_wrap(~status, ncol = 1, scales = "free_y", strip.position = "right") +
  theme_classic()
```

```{r eval = TRUE}
# 6. Table of median log mass by status

# Base R
median.log10.mass <- tapply(mammals$logmass, mammals$status, median, na.rm=TRUE)
cbind(median.log10.mass)

# Tidyverse
mytab <- summarize(group_by(mammals, status), meanlogMass = median(logmass, na.rm = TRUE))
mytab
```

[//]: -->

<br>

### Paired data

Paired data, such as repeated measures, require special plotting techniques to visualize the patterns while preserving information on pairing.

Here we construct a paired data plot for the frequency of extant and extinct mammal species on each continent. 

1. Create a new variable in the mammal data frame that combines the extinct and historical status categories into a single category called "gone". Remove the Insular and Oceanic continents from the data set so we are working only with the continents. Finally, drop the introduction category. To get us more quickly to the plotting step, here's some code to make the new data set:\
<br>Keep only actual continents and drop the "introduction" category from status.
<pre class="fakechunk">
newmammals <- subset(mammals, continent != "Insular" & continent != "Oceanic" & 
                    (status == "extinct" | status == "extant" | status == "historical")  )
</pre>
Make a new status variable that combines the two extinction categories.
<pre class="fakechunk">
newmammals$newstatus <- newmammals$status # initialize the new variable
newmammals$newstatus[newmammals$status %in% c("extinct", "historical")] <- "gone"
</pre>
Make a frequency table of the number of species by continent and newstatus.
<pre class="fakechunk">
mytab <- table(newmammals$continent, newmammals$newstatus, 
                    dnn = c("continent", "newstatus"))
</pre>
We want it in long format, so convert to a "flat" frequency table.
<pre class="fakechunk">
mydata <- data.frame(mytab)
</pre>
Notice it is in long format, which R prefers.
<pre class="fakechunk">
mydata
</pre>

2. Make a paired data plot showing the log10 frequency of species for the two categories of the newstatus variable. What pattern do you notice?

[//]: <!--
### Answers

There is a tendency for the continents with the fewest mammal species to have relatively more species that have lately disappeared. 

```{r eval = TRUE, message = FALSE, include = FALSE}
# This code should be evaluated but is hidden
newmammals <- subset(mammals, continent != "Insular" & continent != "Oceanic" & 
                    (status == "extinct" | status == "extant" | status == "historical")  )

# Make a new status variable that combines the two extinction categories
newmammals$newstatus <- newmammals$status # initialize the new variable
newmammals$newstatus[newmammals$status %in% c("extinct", "historical")] <- "gone"

# Make a frequency table of the number of species by continent and newstatus
mytab <- table(newmammals$continent, newmammals$newstatus, dnn = c("continent", "newstatus"))

# We want it in long format, so convert to a "flat" frequency table.
mydata <- data.frame(mytab)
```

```{r eval = TRUE, message = FALSE}
# 1. The data set you made
print(mydata)

# 2. Paired data plot in base R
interaction.plot(response = log10(mydata$Freq), x.factor = mydata$newstatus, trace.factor = mydata$continent,
    legend = FALSE, lty = 1, xlab = "", ylab = "log10 frequency", 
    type = "b", pch = 16, las = 1, col = "firebrick")
```
    
```{r eval = TRUE, fig.width = 5, fig.height=4}
# Or use ggplot

ggplot(mydata, aes(y = log10(Freq), x = newstatus)) +  
    geom_line(aes(group = continent)) +
    geom_point(size = 3, col = "firebrick", alpha = 0.5) + 
    labs(x = "", y = "log10 frequency") +
    theme_classic() + 
    theme(text = element_text(size = 14))
```

[//]: -->

***

## Data set 2: Fly sex and longevity
The data are from L. Partridge and M. Farquhar (1981), Sexual activity and the lifespan of male fruitflies, <it>Nature</it> 294: 580-581. The experiment placed male fruit flies with varying numbers of previously-mated or virgin females to investigate how mating activity affects male lifespan. The data are in the file fruitflies.csv file on our server [here](https://www.zoology.ubc.ca/~bio501/R/data/fruitflies.csv).

<br>

### Download and inspect
Download the file to your computer and open in a spreadsheet program to have a look at it. View the first few lines of the data frame on the screen, and familiarize yourself with the variable names. 

Our goal here is to find a plot type that clearly and efficiently visualizes the patterns in the data, especially the differences among groups.

<br>

### Analyze

1. Read the data file into a new data frame.

2. Use a strip chart to examine the distribution of longevities in the treatment groups. Try the jitter method to reduce overlap between points. If needed, adjust the size or rotation of the treatment labels so that they all fit on the graph. What pattern of differences between treatments in longevity is revealed? Males from which treatments have the highest longevity? Which have the lowest longevity?

3. Compare the strip chart to a box plot of the same data. Is the pattern in the data as clear in both types of plot?

4. The variable <code>thorax</code> stands for thorax length, which was used as a measure of body size. The measurement was included in case body size also affected longevity. Using ```ggplot()```, produce a scatter plot of thorax length and longevity. Make <code>longevity</code> the response variable (i.e., plot it on the vertical axis). Is there a relationship?

5. Redraw the scatter plot using ```ggplot()``` but this time use different symbols and/or colors for the different treatment groups. Add a legend to identify the symbols. After controlling for differences among males in size, males from which treatments have the highest longevity on average? Which have the lowest longevity?

6. Redraw, adding regression lines to your figure, separately for each group. 

7. You can see how it can be fiendishly difficult to build a clean visual showing the pattern in the data when there are multiple groups. Redraw the figure again using just one color and symbol, but this time use ```facet_wrap()``` to plot the data in multiple panels, one per treatment group. Compare your results with those from (6). Which method shows the differences between groups most clearly? 

[//]: <!--
### Answers

```{r eval = TRUE}
# 1. Read nd inspect data
x <- read.csv(url("https://www.zoology.ubc.ca/~bio501/R/data/fruitflies.csv"),
              stringsAsFactors = FALSE)
head(x)
```

2. Strip chart in base R.

```{r eval = TRUE}
# base R
stripchart(longevity.days ~ treatment, data = x, vertical = TRUE,  
				method = "jitter", pch=16, col = "firebrick", cex.axis=0.7, 
				ylab="Longevity (days)")
```

or using ```ggplot()```

```{r eval = TRUE, fig.width = 5, fig.height=4}
# Using ggplot
ggplot(x, aes(x = treatment, y = longevity.days)) +
    geom_jitter(color = "firebrick", size = 3, width = 0.15) +
    labs(x = "Treatment", y = "Longevity (days)") + 
    theme_classic()
```

3. Box plot. 

```{r eval = TRUE}
boxplot(longevity.days ~ treatment, data = x, cex.axis = 0.7, 
    ylab = "Longevity (days)", boxwex = 0.5, col = "goldenrod1")

```

or using ```ggplot()```

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(x, aes(x = treatment, y = longevity.days)) +
    geom_boxplot(fill = "goldenrod1", width = 0.5) +
    labs(x = "Treatment", y = "Longevity (days)") + 
    theme_classic()
```

4. Scatter plot

```{r eval = TRUE}
# base R
plot(longevity.days ~ thorax.mm, data = x, pch = 16, col = "firebrick", las = 1,
     xlab = "Thorax length (mm)", ylab = "Longevity (days)")
```

or using ```ggplot()```

```{r eval = TRUE, fig.width = 5, fig.height=4}
ggplot(x, aes(x = thorax.mm, y = longevity.days)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Thorax length (mm)", y = "Longevity (days)") + 
    theme_classic()
```

5. Scatter plot with separate colors for each group using ```ggplot()```

```{r eval = TRUE, fig.width = 7, fig.height=4}
ggplot(x, aes(x = thorax.mm, y = longevity.days, colour = treatment, 
            shape = treatment)) + 
	geom_point(size = 2) + 
	labs(x = "Thorax length (mm)", y = "Longevity (days)") + 
	theme_classic()
```

Here's how you would draw a scatter plot with separate colors and symbols for each group using base R. After executing, touch the plot frame with the cursor.

```{r eval = FALSE}
plot(longevity.days ~ thorax.mm, data=x, pch=as.numeric(factor(treatment)), 
        col = as.numeric(factor(treatment)), las = 1, 
        xlab = "Thorax length (mm)", ylab = "Longevity (days)")
legend( locator(1), legend = as.character(levels(factor(x$treatment))),
        pch=1:length(levels(factor(x$treatment))), 
        col=1:length(levels(factor(x$treatment))) )
```

```{r eval = TRUE, fig.width = 7, fig.height=4}
# 6. Add lines using ggplot()
ggplot(x, aes(x=thorax.mm, y=longevity.days, colour = treatment, 
            shape = treatment)) + 
	geom_point(size = 2) +
	geom_smooth(method = lm, linewidth = 1, se = FALSE) +
	labs(x = "Thorax length (mm)", y = "Longevity (days)") + 
	theme_classic()
```

7. Using ```facet_wrap()``` to plot treatment groups in separate panels.

```{r eval = TRUE, fig.width = 7, fig.height=4}
ggplot(x, aes(x = thorax.mm, y = longevity.days)) + 
  facet_wrap(~ treatment, strip.position = "top") +
	geom_point(size = 2) +
	geom_smooth(method = lm, linewidth = 1, se = FALSE) +
	labs(x = "Thorax length (mm)", y = "Longevity (days)") + 
	theme_classic()
```


[//]: -->

\
* <small>
In this mammal data set, the symbol "NA" is used to indicate North America in the ```continent``` column. But in R, NA means missing data. To read properly, you will need to add an argument to your ```read.csv()``` command that changes the missing data code: <code>na.strings=""</code> (or <code>na=""</code> if you are using <code>read_csv()</code> from the <code>readr</code> package). </small>